---
title: "Using IntegratedLearner for Multi-omics Prediction and Classification"
author: "Nalin Arora, Anupreet Porwal and Himel Mallick"
date: "2026-03-10"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{IntegratedLearner}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE, warning=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

java_major_version <- function() {
  out <- tryCatch(
    suppressWarnings(system2("java", "-version", stdout = TRUE, stderr = TRUE)),
    error = function(e) character()
  )
  line1 <- out[1]
  if (length(line1) == 0 || is.na(line1)) return(NA_integer_)
  m <- regmatches(line1, regexec('"([0-9]+)(\\.[0-9]+)?', line1))[[1]]
  if (length(m) < 2) return(NA_integer_)
  as.integer(m[2])
}

has_java_vector_module <- function() {
  out <- tryCatch(
    suppressWarnings(system2("java", "--list-modules", stdout = TRUE, stderr = TRUE)),
    error = function(e) character()
  )
  any(grepl("^jdk.incubator.vector@", out))
}

# Set Java flags before any potential bartMachine/rJava namespace load.
java_params <- "-Xmx6G"
if (has_java_vector_module()) {
  java_params <- c(java_params, "--add-modules=jdk.incubator.vector")
}
options(java.parameters = java_params)

can_run_sl_bart <- function() {
  have_bart <- requireNamespace("bartMachine", quietly = TRUE)
  jv <- java_major_version()
  have_bart && !is.na(jv) && jv >= 21 && has_java_vector_module()
}

use_sl_bart <- isTRUE(can_run_sl_bart())
```

This vignette is a practical tutorial for binary, multiclass, continuous, and survival outcome workflows in `IntegratedLearner`.

The goal is to show a complete end-to-end pattern you can adapt to your own multi-omics study:

1. Build correctly formatted training/validation inputs.
2. Fit per-layer, stacked, and concatenated learners.
3. Interpret model outputs (AUC/R2, layer weights, and feature signals).

`IntegratedLearner` supports three integration paradigms:

- Early fusion: concatenated features across layers.
- Late fusion: layer-specific models combined by a meta-learner.
- Intermediate fusion: additional integration strategies for downstream analysis.

## Load Packages

```{r, warning=FALSE, message=FALSE}
# Main package
devtools::install_github("himelmallick/IntegratedLearner")
library(IntegratedLearner)

# Tutorial dependencies
library(tidyverse)
library(SuperLearner)
library(caret)
library(cowplot)
library(bayesplot)
library(curatedMetagenomicData)
library(TreeSummarizedExperiment)
library(SummarizedExperiment)
library(MultiAssayExperiment)
library(survival)
library(S4Vectors)
if (use_sl_bart) {
  library(bartMachine)
}
```

```{r, include=FALSE}
if (!use_sl_bart) {
  message(
    "SL.BART sections are in fallback mode: Java >= 21 with bartMachine is required for SL.BART."
  )
}
```

## Input Data Contract

For the `PCL_*` interface used in this tutorial, each dataset is a list with:

- `feature_table`: data frame with features in rows and samples in columns.
- `sample_metadata`: data frame with samples in rows. Must include:
  - `subjectID`: subject identifier (can repeat for longitudinal data).
  - `Y`: outcome (binary, continuous, or survival-encoded).
- `feature_metadata`: data frame with features in rows. Must include:
  - `featureID`: unique feature identifier.
  - `featureType`: layer label (for example, species, metabolites).

Required alignments:

- `rownames(feature_table) == rownames(feature_metadata)`
- `colnames(feature_table) == rownames(sample_metadata)`

If you provide a validation set, it must use the same feature set and ordering as training.

For survival workflows, include `time` and `event` columns in `sample_metadata` (with `event` coded as 0/1). You can also provide `Y` as a `Surv(time, event)` object.

## Alternative Input Mode: MAE (Complete Binary Example)

`IntegratedLearner` accepts `MultiAssayExperiment` inputs through `MAE_train`/`MAE_valid`. This is often the cleanest path when each omics layer is already represented as a `SummarizedExperiment`/`TreeSummarizedExperiment`.


```{r, eval=TRUE,warning = FALSE, eval=FALSE}


# 1) Download two aligned layers from curatedMetagenomicData
asnicar_tax <- curatedMetagenomicData(
  "DavidLA_2015.relative_abundance",
  dryrun = FALSE
)[[1]]

asnicar_path <- curatedMetagenomicData(
  "DavidLA_2015.pathway_abundance",
  dryrun = FALSE
)[[1]]

tax_tse  <- as(asnicar_tax,  "TreeSummarizedExperiment")
path_tse <- as(asnicar_path, "TreeSummarizedExperiment")

# 2) Keep common samples in both layers
common_samples <- intersect(colnames(tax_tse), colnames(path_tse))
common_samples <- as.character(common_samples)

tax_tse  <- tax_tse[, common_samples]
path_tse <- path_tse[, common_samples]

# 3) Build binary outcome and subject IDs inside each experiment
Yvec <- ifelse(as.character(colData(tax_tse)$disease) == "healthy", 0L, 1L)

SummarizedExperiment::colData(tax_tse)$Y <- Yvec
SummarizedExperiment::colData(path_tse)$Y <- Yvec

SummarizedExperiment::colData(tax_tse)$subjectID <- common_samples
SummarizedExperiment::colData(path_tse)$subjectID <- common_samples

# 4) Build top-level MAE colData
cd <- S4Vectors::DataFrame(
  Y = as.integer(Yvec),
  subjectID = common_samples,
  row.names = common_samples
)

# 5) Build explicit sampleMap
smap <- S4Vectors::DataFrame(
  assay   = c(rep("taxonomy", length(common_samples)),
              rep("pathway",  length(common_samples))),
  primary = c(common_samples, common_samples),
  colname = c(common_samples, common_samples)
)

smap$assay   <- as.character(smap$assay)
smap$primary <- as.character(smap$primary)
smap$colname <- as.character(smap$colname)

# 6) Build MAE container
mae <- MultiAssayExperiment(
  experiments = ExperimentList(
    taxonomy = tax_tse,
    pathway  = path_tse
  ),
  colData = cd,
  sampleMap = smap
)

# 7) Stratified train/validation split
y <- MultiAssayExperiment::colData(mae)$Y
names(y) <- rownames(MultiAssayExperiment::colData(mae))

set.seed(1)

i0 <- which(y == 0)
i1 <- which(y == 1)

train0 <- sample(i0, floor(0.7 * length(i0)))
train1 <- sample(i1, floor(0.7 * length(i1)))

train_ids <- names(y)[sort(c(train0, train1))]
valid_ids <- setdiff(names(y), train_ids)

mae_train <- mae[, train_ids]
mae_valid <- mae[, valid_ids]

# 8) Fit IntegratedLearner in MAE mode
fit_mae_bin <- IntegratedLearner(
  MAE_train = mae_train,
  MAE_valid = mae_valid,
  experiment = c("taxonomy", "pathway"),
  assay.type = c("relative_abundance", "pathway_abundance"),
  folds = 2,
  base_learner = "SL.randomForest",
  meta_learner = "SL.nnls.auc",
  family = binomial(),
  verbose = TRUE
)

# 10) Results
fit_mae_bin$AUC.train
fit_mae_bin$AUC.test
fit_mae_bin$weights
```

The returned object is the same style as `PCL` mode, so downstream interpretation (`AUC.train`, `R2.train`, `weights`, `plot.learner`) is unchanged.

## Parameter Reference (Conbin, Multiclass, and Survival)

### Common Wrapper Parameters (`IntegratedLearner`)

| Parameter | Default | Applies to | Description |
|---|---|---|---|
| `MAE_train`, `MAE_valid` | `NULL` | all | MAE-mode inputs (training and optional validation). |
| `PCL_train`, `PCL_valid` | `NULL` | all | PCL-mode inputs (training and optional validation). |
| `experiment` | `NULL` | MAE mode | Selected MAE experiment names/indices; defaults to all experiments. |
| `assay.type` | `NULL` | MAE mode | Assay names per selected experiment. |
| `na.rm` | `FALSE` | all | Drop features with missing values after extraction/prep. |
| `folds` | `5` | all | Outer CV folds. |
| `seed` | `1234` | all | Reproducibility seed. |
| `base_learner` | `"SL.BART"` | all | Base learner. Use `SL.*` IDs for continuous/binary, native multiclass IDs (for example `randomforest`, `xgboost`, `mbart`) for multiclass, and explicitly set a supported `surv.*` ID for survival runs. |
| `verbose` | `FALSE` | all | Print progress. |
| `family` | `gaussian()` | all | Non-survival: `gaussian()`/`binomial()`. Multiclass is auto-detected when `family = binomial()` and `Y` has more than two classes. Survival is auto-detected from metadata or family. |
| `...` | — | all | Passed to the selected backend (`IL_conbin` or `ILsurv`). |

### Conbin-Specific Parameters (Continuous/Binary Path)

| Parameter | Default | Description |
|---|---|---|
| `base_screener` | `"All"` | SuperLearner screener used with base learner. |
| `meta_learner` | `"SL.nnls.auc"` | Stacked meta learner for late fusion. |
| `run_stacked` | `TRUE` | Enables late-fusion stacked model. |
| `run_concat` | `TRUE` | Enables early-fusion concatenated model. |
| `print_learner` | `TRUE` | Prints fit summary. |
| `refit.stack` | `FALSE` | Refit stacked learner on full data for final predictions. |

### Multiclass-Specific Parameters (Native Multiclass Path)

| Parameter | Default | Description |
|---|---|---|
| `base_learner` | `"glmnet"` | Native multiclass learner per layer and for concatenated fit. Supported: `glmnet`, `randomforest`, `ranger`, `xgboost`, `mbart`, `multinom`. |
| `meta_learner` | `"glmnet"` | Native multiclass learner used for stacked fusion. |
| `base_screener` | `"All"` | Optional multiclass feature screening before layer/concat model fit (`All`, `anova`, `glmnet`, `randomforest`, `ranger`, `xgboost`). |
| `run_stacked` | `TRUE` | Enables late-fusion stacked multiclass model. |
| `run_concat` | `TRUE` | Enables early-fusion concatenated multiclass model. |
| `folds` | `5` | Subject-level CV folds for OOF multiclass probabilities. |
| `...` (`screen_*`) | — | Screener controls (for example `screen_keep_n`, `screen_keep_prop`, `screen_max_features`, `screen_min_features`). |

### Survival-Specific Parameters (via `...`)

| Parameter | Default | Description |
|---|---|---|
| `do_early_fusion` | `TRUE` | Train an early-fusion survival model on all features. |
| `weight_method` | `"IBS"` | Late-fusion weighting objective (`"IBS"` or `"COX"`). |
| `intermediate_learners` | `c("surv.coxph")` | Intermediate fusion learners on layer-risk features. |
| `t_vec`, `t_vec_probs` | `NULL`, quantiles | Time grid used in COX-style weighting summaries. |
| `layer_score` | `"sum"` | Aggregation of cumulative hazard increments (`sum`, `mean`, `l2`). |
| `weight_lambda` | `0.02` | Regularization strength for COX weighting optimizer. |
| `weight_penalty` | `"l2_to_uniform"` or `"entropy"` | Penalty used while learning survival late-fusion weights. |
| `weight_cap` | `1.0` | Optional cap on individual layer weights. |
| `optim_maxit_cox` | `4000` | Max iterations for COX weighting optimization. |
| `optim_maxit_ibs` | `300` | Max iterations for IBS weighting optimization. |
| `ibs_shrink_to_uniform` | `0` | Shrink IBS weights toward uniform blend. |

## Supported Models and Fusion Modules

### Supported Models

| Path | Supported base models |
|---|---|
| Continuous/Binary (`IL_conbin`) | Any `SuperLearner`-compatible `SL.*` learner available in your R session. Package wrappers include: `SL.BART`, `SL.LASSO`, `SL.enet`, `SL.glmnet2`, `SL.horseshoe`, `SL.mxBART` (plus standard SuperLearner learners such as `SL.glm`, `SL.randomForest`, etc.). |
| Multiclass (`IL_multiclass`) | Native multiclass learner IDs: `glmnet`, `randomforest`, `ranger`, `xgboost`, `mbart`, `multinom`. |
| Survival (`ILsurv`) | Built-in survival learner IDs: `surv.coxph`, `surv.glmnet`, `surv.ranger`, `surv.ranger.extratrees`, `surv.ranger.maxstat`, `surv.ranger.C`, `surv.rfsrc`, `surv.coxboost`, `surv.gbm`, `surv.xgboost.cox`, `surv.xgboost.aft`, `surv.mboost`, `surv.bart`. |

### Supported Fusion Outputs

| Path | Single-layer | Early fusion | Late fusion | Intermediate fusion |
|---|---|---|---|---|
| Continuous/Binary | Yes | `run_concat = TRUE` | `run_stacked = TRUE` with `meta_learner` | Not a separate branch in `IL_conbin` |
| Multiclass | Yes | `run_concat = TRUE` | `run_stacked = TRUE` with native multiclass `meta_learner` | Not a separate branch in `IL_multiclass` |
| Survival | Yes (`train_out$single`) | `do_early_fusion = TRUE` | Weighted layer blending (`weight_method = "IBS"` or `"COX"`) | `intermediate_learners` on layer-risk summaries |

## Output Reference: What You Get and How to Access It

This section summarizes the outputs produced by each integration method and where to find weights/importance values.

### Conbin Outputs (Binary/Continuous)

| Method | What it returns | Where to access |
|---|---|---|
| Single-layer (per omics layer) | Layer-specific predictions and metrics | `fit$yhat.train[, layer_name]`, `fit$yhat.test[, layer_name]` (if validation), `fit$AUC.train` / `fit$AUC.test` (binomial), `fit$R2.train` / `fit$R2.test` (gaussian) |
| Early fusion (concatenated) | One model on all features concatenated | Enable with `run_concat = TRUE`; outputs in `fit$yhat.train[, "concatenated"]`, `fit$model_fits$model_concat`, `fit$SL_fits$SL_fit_concat` |
| Late fusion (stacked) | Meta-model over layer-level predictions | Enable with `run_stacked = TRUE`; outputs in `fit$yhat.train[, "stacked"]`, `fit$model_fits$model_stacked`, `fit$SL_fits$SL_fit_stacked` |
| Layer weights (stacked) | Relative contribution of each layer in late fusion | `fit$weights` (available when `meta_learner = "SL.nnls.auc"` and `run_stacked = TRUE`) |

### Multiclass Outputs

| Method | What it returns | Where to access |
|---|---|---|
| Single-layer (per omics layer) | Layer-wise multiclass probability and class predictions | `fit$prob.train[[layer_name]]`, `fit$class.train[, layer_name]`, plus validation analogs `fit$prob.test[[layer_name]]`, `fit$class.test[, layer_name]` |
| Early fusion (concatenated) | One multiclass model on concatenated features | Enable with `run_concat = TRUE`; outputs in `fit$prob.train$concatenated`, `fit$class.train[, "concatenated"]`, `fit$model_fits$model_concat` |
| Late fusion (stacked) | Multiclass meta-model over OOF layer probabilities | Enable with `run_stacked = TRUE`; outputs in `fit$prob.train$stacked`, `fit$class.train[, "stacked"]`, `fit$model_fits$model_stacked` |
| Multiclass performance metrics | Accuracy, balanced accuracy, and log-loss | `fit$metrics.train` and `fit$metrics.test` (if validation provided) |
| Screened feature sets | Features retained by multiclass screeners | `fit$selected_features_by_layer`, `fit$selected_features_concat` |

### Survival Outputs (Single/Early/Late/Intermediate)

| Method | Training outputs | Validation outputs |
|---|---|---|
| Single-layer | `fit$train_out$single$metrics` | `fit$valid_out$single$valid_cindex`, `fit$valid_out$single$valid_auc` |
| Early fusion | `fit$train_out$early$train_cindex`, `fit$train_out$early$train_auc` | `fit$valid_out$early$valid_cindex`, `fit$valid_out$early$valid_auc` |
| Late fusion | `fit$train_out$late$weights`, `fit$train_out$late$train_cindex`, `fit$train_out$late$train_auc` | `fit$valid_out$late$valid_cindex`, `fit$valid_out$late$valid_auc` |
| Intermediate fusion | `fit$train_out$intermediate` (per learner listed in `intermediate_learners`) | `fit$valid_out$intermediate` |

### Importance Outputs (Conbin, Multiclass, and Survival)

| Importance type | Where to access | Notes |
|---|---|---|
| Conbin signed global feature importance | `fit$feature_importance_signed` | Always returned for non-survival fits; named numeric vector sorted by effect magnitude/sign. |
| Conbin signed per-layer importance | `fit$feature_importance_signed_by_layer` | List split by `featureType`. |
| Multiclass signed global feature importance | `fit$feature_importance_global` | Global score aggregated across multiclass contrasts. |
| Multiclass signed importance by class | `fit$feature_importance_signed_by_class` | List with one signed vector per class. |
| Multiclass signed importance by layer and class | `fit$feature_importance_signed_by_layer_by_class` | Nested list by layer then class. |
| Survival early-fusion combined importance | `fit$train_out$early$combined_importance` | Available when `do_early_fusion = TRUE`. |
| Survival late-fusion combined importance | `fit$train_out$late$combined_importance` | Weighted signed importance; names are prefixed like `layer::feature`. |
| BART-specific layer importance (optional) | `bartMachine::investigate_var_importance(fit$model_fits$model_layers[[layer]], plot = FALSE)` | Only for BART-based conbin fits (`base_learner = "SL.BART"`). |

### Quick Access Snippets

```{r, eval=FALSE}
# ---- Conbin: weights + top features ----
fit$weights
head(fit$feature_importance_signed, 20)
names(fit$feature_importance_signed_by_layer)
head(fit$feature_importance_signed_by_layer[[1]], 20)

# ---- Multiclass: metrics + class probabilities + importance ----
fit_mc$metrics.train
fit_mc$metrics.test
head(fit_mc$class.train)
head(fit_mc$class.test)
head(fit_mc$prob.train$stacked)
head(fit_mc$feature_importance_global, 20)
head(fit_mc$feature_importance_signed_by_class[[1]], 20)

# ---- Survival: late-fusion weights + top combined features ----
fit_surv$train_out$late$weights
head(fit_surv$train_out$late$combined_importance, 20)

# ---- Survival: inspect all fusion branches ----
fit_surv$train_out$single
fit_surv$train_out$early
fit_surv$train_out$late
fit_surv$train_out$intermediate
```

## Example 1: Binary Outcome (IBD Classification)

This section uses the PRISM dataset (Franzosa et al., 2019) for classifying IBD status. The binary target is coded in `sample_metadata$Y`.

### Step 1: Load and Inspect Training and Validation Data

```{r}
# Training data
load(url("https://github.com/himelmallick/IntegratedLearner/blob/master/data/PRISM.RData?raw=true"))

feature_table <- pcl$feature_table
sample_metadata <- pcl$sample_metadata
feature_metadata <- pcl$feature_metadata
rm(pcl)

# Quick checks
head(feature_table[1:5, 1:5])
head(sample_metadata[1:5, ])
head(feature_metadata[1:5, ])

table(feature_metadata$featureType)
table(sample_metadata$Y)

all(rownames(feature_table) == rownames(feature_metadata))
all(colnames(feature_table) == rownames(sample_metadata))

# Independent validation data
load(url("https://github.com/himelmallick/IntegratedLearner/blob/master/data/NLIBD.RData?raw=true"))
feature_table_valid <- pcl$feature_table
sample_metadata_valid <- pcl$sample_metadata
rm(pcl)

all(rownames(feature_table) == rownames(feature_table_valid))
all(colnames(feature_table_valid) == rownames(sample_metadata_valid))
```

### Step 2: Build PCL Inputs

```{r}
PCL_train <- list(
  feature_table = feature_table,
  sample_metadata = sample_metadata,
  feature_metadata = feature_metadata
)

PCL_valid <- list(
  feature_table = feature_table_valid,
  sample_metadata = sample_metadata_valid,
  feature_metadata = feature_metadata
)
```

### Step 3: Fit the Model

`IntegratedLearner` fits one model per layer (`base_learner`) and then combines layer-level predictions with a meta-learner (`meta_learner`).

```{r, warning=FALSE}
fit <- IntegratedLearner(
  PCL_train = PCL_train,
  PCL_valid = PCL_valid,
  folds = 2,
  base_learner = "SL.randomForest",
  meta_learner = "SL.nnls.auc",
  verbose = TRUE,
  family = binomial()
)
```

### Step 4: Inspect and Interpret Outputs

Core outputs for binary tasks include:

- `fit$AUC.train` and `fit$AUC.test`: AUC per layer and fusion model.
- `fit$weights`: layer contributions in the stacked model (when `SL.nnls.auc` is used).
- `fit$yhat.train` and `fit$yhat.test`: predicted probabilities.

```{r}
fit$AUC.train
fit$AUC.test
fit$weights
```

Plot ROC summaries for train and validation sets:

```{r, warning=FALSE, fig.height=10, fig.width=7}
plot.obj <- IntegratedLearner:::plot.learner(fit)
```

In this PRISM setting, you can compare which single layer is strongest and whether stacked fusion outperforms both individual layers and simple concatenation.

## Example 2: Continuous Outcome (Gestational Age)

This section uses the pregnancy dataset (Ghaemi et al., 2019), where `Y` is continuous gestational age.

### Step 1: Load and Inspect Data

```{r}
load(url("https://github.com/himelmallick/IntegratedLearner/blob/master/data/pregnancy.RData?raw=true"))

feature_table <- pcl$feature_table
sample_metadata <- pcl$sample_metadata
feature_metadata <- pcl$feature_metadata
rm(pcl)

head(feature_table[1:5, 1:5])
head(sample_metadata[1:5, ])
head(feature_metadata[1:5, ])

table(feature_metadata$featureType)
length(unique(sample_metadata$subjectID))

all(rownames(feature_table) == rownames(feature_metadata))
all(colnames(feature_table) == rownames(sample_metadata))

# Optional speed-up for local experimentation
# top_n <- 50
# subsetIDs <- c(1:top_n, (nrow(feature_table) - top_n + 1):nrow(feature_table))
# feature_table <- feature_table[subsetIDs, ]
# feature_metadata <- feature_metadata[subsetIDs, ]
```

### Step 2: Build PCL Input

```{r}
PCL_train <- list(
  feature_table = feature_table,
  sample_metadata = sample_metadata,
  feature_metadata = feature_metadata
)
```

### Step 3: Fit Continuous Model

For this example, we use BART base learners (`SL.BART`).

If you hit:

`java.lang.UnsupportedClassVersionError ... class file version 65.0 ... recognizes up to 61.0`

your Java runtime is older than the version used by your installed `bartMachine` build (typically Java 17 runtime vs Java 21 bytecode). In that case, either:

- upgrade Java runtime to 21 and restart R, or
- use a non-Java learner (example fallback shown below).

```{r, warning=FALSE, eval=use_sl_bart}
fit <- IntegratedLearner(
  PCL_train = PCL_train,
  folds = 2,
  base_learner = "SL.BART",
  meta_learner = "SL.nnls.auc",
  family = gaussian()
)
```

Fallback (non-Java) run:

```{r, eval=!use_sl_bart}
fit <- IntegratedLearner(
  PCL_train = PCL_train,
  folds = 2,
  base_learner = "SL.randomForest",
  meta_learner = "SL.nnls.auc",
  family = gaussian()
)
```

### Step 4: Evaluate Predictive Accuracy

For continuous outcomes, `IntegratedLearner` reports `R2.train` (and `R2.test` if validation is provided).

```{r}
fit$R2.train
```

```{r fig.asp=0.8, fig.width=7}
plot.obj <- IntegratedLearner:::plot.learner(fit)
```

### Step 5: Uncertainty and Feature-Level Interpretation (BART)

When using `SL.BART`, you can inspect posterior predictive distributions and derive weighted posterior summaries.
`r if (!use_sl_bart) "This section is skipped in this session because Java < 21 (or bartMachine is unavailable)." else ""`

```{r, eval=use_sl_bart}
weights <- fit$weights

dataX <- fit$X_train_layers
dataY <- fit$Y_train

post.samples <- vector("list", length(weights))
names(post.samples) <- names(dataX)

for (i in seq_along(post.samples)) {
  post.samples[[i]] <- bartMachine::bart_machine_get_posterior(
    fit$model_fits$model_layers[[i]],
    dataX[[i]]
  )$y_hat_posterior_samples
}

weighted.post.samples <- Reduce("+", Map("*", post.samples, weights))
rownames(weighted.post.samples) <- rownames(dataX[[1]])
names(dataY) <- rownames(dataX[[1]])
```

Visualize 68% and 95% credible intervals for observations:

```{r fig.width=9, fig.height=7, eval=use_sl_bart}
ord_names <- names(sort(rowMeans(weighted.post.samples), decreasing = TRUE))

mcmc_intervals(t(weighted.post.samples), prob = 0.68, prob_outer = 0.95) +
  scale_y_discrete(limits = ord_names) +
  geom_point(aes(x = dataY[ord_names], y = ord_names), shape = 1, size = 3, color = "black") +
  coord_flip() +
  theme_bw() +
  labs(
    x = "Gestational age (in months)",
    y = "Observations"
  ) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

Layer weights and feature-level inclusion proportions can also be examined for biological interpretation.

```{r, eval=use_sl_bart}
omicsEye_theme <- function() {
  angle <- 45
  hjust <- 1
  ggplot2::theme_bw() +
    ggplot2::theme(
      axis.text.x = ggplot2::element_text(size = 8, vjust = 1, hjust = hjust, angle = angle),
      axis.text.y = ggplot2::element_text(size = 8, hjust = 1),
      axis.title = ggplot2::element_text(size = 10),
      plot.title = ggplot2::element_text(size = 10),
      plot.subtitle = ggplot2::element_text(size = 8),
      legend.title = ggplot2::element_text(size = 6, face = "bold"),
      legend.text = ggplot2::element_text(size = 7),
      axis.line = ggplot2::element_line(colour = "black", linewidth = 0.25),
      axis.line.x = ggplot2::element_line(colour = "black", linewidth = 0.25),
      axis.line.y = ggplot2::element_line(colour = "black", linewidth = 0.25),
      panel.border = ggplot2::element_blank(),
      panel.grid.major = ggplot2::element_blank(),
      panel.grid.minor = ggplot2::element_blank()
    )
}

safe_var_importance <- function(model, layer_label) {
  tryCatch({
    qq <- bartMachine::investigate_var_importance(model, plot = FALSE)
    df <- cbind.data.frame(qq$avg_var_props, qq$sd_var_props)
    colnames(df) <- c("mean", "sd")
    df$type <- layer_label
    df
  }, error = function(e) {
    warning(sprintf("Skipping variable importance for %s: %s", layer_label, conditionMessage(e)))
    data.frame(mean = numeric(), sd = numeric(), type = character())
  })
}

vimp_stack <- cbind.data.frame(fit$weights)
colnames(vimp_stack) <- "mean"
vimp_stack$sd <- NA
vimp_stack$type <- "stack"

layer_names <- names(fit$model_fits$model_layers)
vimp_layers <- lapply(layer_names, function(layer_nm) {
  safe_var_importance(fit$model_fits$model_layers[[layer_nm]], layer_nm)
})

vimp_layers <- vimp_layers[lengths(vimp_layers) > 0]
vimp_top <- do.call(
  rbind,
  lapply(vimp_layers, function(df) head(df[order(-df$mean), , drop = FALSE], 20))
)

VIMP <- as.data.frame(rbind.data.frame(vimp_stack, vimp_top))
VIMP <- tibble::rownames_to_column(VIMP, "ID")

p4 <- VIMP %>%
  dplyr::filter(type == "stack") %>%
  dplyr::arrange(desc(mean)) %>%
  ggplot(aes(y = mean, x = reorder(ID, -mean))) +
  geom_bar(stat = "identity", fill = "darkseagreen") +
  theme_bw() +
  omicsEye_theme() +
  ylab("Layer Weights") +
  xlab("")

p5 <- VIMP %>%
  dplyr::filter(type != "stack") %>%
  dplyr::arrange(mean) %>%
  dplyr::mutate(ID = stringr::str_replace_all(ID, stringr::fixed("_"), " ")) %>%
  ggplot(aes(reorder(ID, -mean), mean, fill = type)) +
  facet_wrap(. ~ type, scales = "free") +
  geom_bar(stat = "identity", fill = "lightsalmon") +
  geom_errorbar(aes(ymin = ifelse(mean - sd > 0, mean - sd, 0), ymax = mean + sd),
                width = 0.2,
                position = position_dodge(0.9)) +
  theme_bw() +
  coord_flip() +
  omicsEye_theme() +
  theme(strip.background = element_blank()) +
  ylab("Inclusion proportion") +
  xlab("")
```

```{r fig.width=7, fig.height=4, eval=use_sl_bart}
plot_grid(
  p4,
  ncol = 1,
  labels = c("Estimated IntegratedLearner Layer Weights"),
  label_size = 8,
  vjust = 0.1
) + theme(plot.margin = unit(c(0.5, 0.5, 0.5, 0.5), "cm"))
```

```{r fig.width=9, fig.height=7, eval=use_sl_bart}
plot_grid(
  p5,
  ncol = 1,
  labels = c("Top Features by Layer (BART Inclusion Proportions)"),
  label_size = 8,
  vjust = 0.1
) + theme(plot.margin = unit(c(0.5, 0.5, 0.5, 0.5), "cm"))
```

## Example 3: Multiclass Outcome (Franzosa MAE with External Validation)

This section shows a full multiclass MAE workflow using `diseaseCat` as `Y`, with training/validation feature alignment and Python-style class balancing on the training set.

```{r}

load(url("https://github.com/himelmallick/IntegratedLearner/blob/master/data/FranzosaE_2019_Train_Validation.RData?raw=true"))
fit <- IntegratedLearner::IntegratedLearner(
  MAE_train = MAE_train,
  MAE_valid = MAE_valid,
  experiment = c("metabolome", "species"),
  assay.type = c("abundance", "abundance"),
  family = stats::binomial(),
  base_learner = "mbart",
  meta_learner = "mbart",
  run_stacked = TRUE,
  run_concat = TRUE,
  folds = 2,
  verbose = TRUE
)

```

Useful multiclass outputs:

- `fit$metrics.train`
- `fit$metrics.test`
- `fit$class.train`
- `fit$class.test`
- `fit$feature_importance_signed_by_class`

## Example 4: Survival Outcome (Time-to-event)

For survival tasks, `IntegratedLearner` dispatches to `ILsurv` when survival metadata are detected. The expected fields are:

- `time`: follow-up time (non-negative).
- `event`: event indicator (`0/1`).
- optional `Y`: `Surv(time, event)` convenience column.

This path uses the package-native survival backend (no `mlr3` dependency required).

This section provides a complete MAE workflow, followed by an equivalent PCL sketch.



```{r}


load(url("https://github.com/himelmallick/IntegratedLearner/blob/master/data/TCGA_BRCA.RData?raw=true"))
fit_surv_mae <- IntegratedLearner(
  MAE_train = mae_train,
  MAE_valid = mae_valid,
  experiment = c("taxonomy", "pathway"),
  assay.type = c("relative_abundance", "pathway_abundance"),
  folds = 2,
  base_learner = "surv.xgboost.cox",  # alternatives include surv.coxph/surv.ranger/surv.xgboost.cox/...
  weight_method = "COX",           # alternative: "IBS"
  verbose = TRUE
)
```




### Interpret Survival Outputs

```{r}
# Single-layer training metrics
# fit_surv_mae$train_out$single$metrics

# Late fusion (weighted integration)
fit_surv_mae$train_out$late$weights
fit_surv_mae$train_out$late$train_cindex
fit_surv_mae$train_out$late$train_auc

# Early and intermediate fusion summaries
#fit_surv_mae$train_out$early
#fit_surv_mae$train_out$intermediate

# Validation metrics
fit_surv_mae$valid_out$late$valid_cindex
fit_surv_mae$valid_out$late$valid_auc
fit_surv_mae$valid_out$single$valid_cindex
```

### Equivalent PCL Survival 

```{r,warning =FALSE}

fit_surv_pcl <- IntegratedLearner(
  PCL_train = PCL_train,
  PCL_valid = PCL_valid,
  folds = 2,
  base_learner = "surv.coxboost",
  weight_method = "IBS"
)
```

## Session Information

```{r}
sessionInfo()
```

## References

Franzosa EA et al. (2019). [Gut microbiome structure and metabolic activity in inflammatory bowel disease](https://www.ncbi.nlm.nih.gov/pubmed/30531976). *Nature Microbiology* 4(2):293-305.

Ghaemi MS et al. (2019). [Multiomics modeling of the immunome, transcriptome, microbiome, proteome and metabolome adaptations during human pregnancy](https://pubmed.ncbi.nlm.nih.gov/30561547/). *Bioinformatics* 35(1):95-103.

## Citation

Mallick et al. (2024). [An integrated Bayesian framework for multi-omics prediction and classification](https://pubmed.ncbi.nlm.nih.gov/38146838/). *Statistics in Medicine* 43(5):983-1002.
