---
title: "Using IntegratedLearner for Multi-omics Prediction and Classification"
author: "Himel Mallick, Nalin Arora and Anupreet Porwal"
date: "2026-03-10"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{IntegratedLearner}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette is a practical tutorial for binary, continuous, and survival outcome workflows in `IntegratedLearner`.

The goal is to show a complete end-to-end pattern you can adapt to your own multi-omics study:

1. Build correctly formatted training/validation inputs.
2. Fit per-layer, stacked, and concatenated learners.
3. Interpret model outputs (AUC/R2, layer weights, and feature signals).

`IntegratedLearner` supports three integration paradigms:

- Early fusion: concatenated features across layers.
- Late fusion: layer-specific models combined by a meta-learner.
- Intermediate fusion: additional integration strategies for downstream analysis.

## Load Packages

```{r, warning=FALSE, message=FALSE}
# Main package
library(IntegratedLearner)

# Needed for BART-based sections
options(java.parameters = "-Xmx6G")

# Tutorial dependencies
library(tidyverse)
library(SuperLearner)
library(caret)
library(cowplot)
library(bayesplot)
library(bartMachine)
library(curatedMetagenomicData)
library(TreeSummarizedExperiment)
library(SummarizedExperiment)
library(MultiAssayExperiment)
library(mlr3)
library(mlr3proba)
library(mlr3pipelines)
library(mlr3extralearners)
library(survival)
library(S4Vectors)
```

## Input Data Contract

For the `PCL_*` interface used in this tutorial, each dataset is a list with:

- `feature_table`: data frame with features in rows and samples in columns.
- `sample_metadata`: data frame with samples in rows. Must include:
  - `subjectID`: subject identifier (can repeat for longitudinal data).
  - `Y`: outcome (binary, continuous, or survival-encoded).
- `feature_metadata`: data frame with features in rows. Must include:
  - `featureID`: unique feature identifier.
  - `featureType`: layer label (for example, species, metabolites).

Required alignments:

- `rownames(feature_table) == rownames(feature_metadata)`
- `colnames(feature_table) == rownames(sample_metadata)`

If you provide a validation set, it must use the same feature set and ordering as training.

For survival workflows, include `time` and `event` columns in `sample_metadata` (with `event` coded as 0/1). You can also provide `Y` as a `Surv(time, event)` object.

## Alternative Input Mode: MAE (Complete Binary Example)

`IntegratedLearner` accepts `MultiAssayExperiment` inputs through `MAE_train`/`MAE_valid`. This is often the cleanest path when each omics layer is already represented as a `SummarizedExperiment`/`TreeSummarizedExperiment`.


```{r, eval=TRUE,warning = FALSE}


# 1) Download two aligned layers from curatedMetagenomicData
asnicar_tax <- curatedMetagenomicData(
  "DavidLA_2015.relative_abundance",
  dryrun = FALSE
)[[1]]

asnicar_path <- curatedMetagenomicData(
  "DavidLA_2015.pathway_abundance",
  dryrun = FALSE
)[[1]]

tax_tse  <- as(asnicar_tax,  "TreeSummarizedExperiment")
path_tse <- as(asnicar_path, "TreeSummarizedExperiment")

# 2) Keep common samples in both layers
common_samples <- intersect(colnames(tax_tse), colnames(path_tse))
common_samples <- as.character(common_samples)

tax_tse  <- tax_tse[, common_samples]
path_tse <- path_tse[, common_samples]

# 3) Build binary outcome and subject IDs inside each experiment
Yvec <- ifelse(as.character(colData(tax_tse)$disease) == "healthy", 0L, 1L)

SummarizedExperiment::colData(tax_tse)$Y <- Yvec
SummarizedExperiment::colData(path_tse)$Y <- Yvec

SummarizedExperiment::colData(tax_tse)$subjectID <- common_samples
SummarizedExperiment::colData(path_tse)$subjectID <- common_samples

# 4) Build top-level MAE colData
cd <- S4Vectors::DataFrame(
  Y = as.integer(Yvec),
  subjectID = common_samples,
  row.names = common_samples
)

# 5) Build explicit sampleMap
smap <- S4Vectors::DataFrame(
  assay   = c(rep("taxonomy", length(common_samples)),
              rep("pathway",  length(common_samples))),
  primary = c(common_samples, common_samples),
  colname = c(common_samples, common_samples)
)

smap$assay   <- as.character(smap$assay)
smap$primary <- as.character(smap$primary)
smap$colname <- as.character(smap$colname)

# 6) Build MAE container
mae <- MultiAssayExperiment(
  experiments = ExperimentList(
    taxonomy = tax_tse,
    pathway  = path_tse
  ),
  colData = cd,
  sampleMap = smap
)

# 7) Stratified train/validation split
y <- MultiAssayExperiment::colData(mae)$Y
names(y) <- rownames(MultiAssayExperiment::colData(mae))

set.seed(1)

i0 <- which(y == 0)
i1 <- which(y == 1)

train0 <- sample(i0, floor(0.7 * length(i0)))
train1 <- sample(i1, floor(0.7 * length(i1)))

train_ids <- names(y)[sort(c(train0, train1))]
valid_ids <- setdiff(names(y), train_ids)

mae_train <- mae[, train_ids]
mae_valid <- mae[, valid_ids]

# 8) Fit IntegratedLearner in MAE mode
fit_mae_bin <- IntegratedLearner(
  MAE_train = mae_train,
  MAE_valid = mae_valid,
  experiment = c("taxonomy", "pathway"),
  assay.type = c("relative_abundance", "pathway_abundance"),
  folds = 3,
  base_learner = "SL.randomForest",
  meta_learner = "SL.nnls.auc",
  family = binomial(),
  verbose = TRUE
)

# 10) Results
fit_mae_bin$AUC.train
fit_mae_bin$AUC.test
fit_mae_bin$weights
```

The returned object is the same style as `PCL` mode, so downstream interpretation (`AUC.train`, `R2.train`, `weights`, `plot.learner`) is unchanged.

## Parameter Reference (Conbin and Survival)

### Common Wrapper Parameters (`IntegratedLearner`)

| Parameter | Default | Applies to | Description |
|---|---|---|---|
| `MAE_train`, `MAE_valid` | `NULL` | all | MAE-mode inputs (training and optional validation). |
| `PCL_train`, `PCL_valid` | `NULL` | all | PCL-mode inputs (training and optional validation). |
| `experiment` | `NULL` | MAE mode | Selected MAE experiment names/indices; defaults to all experiments. |
| `assay.type` | `NULL` | MAE mode | Assay names per selected experiment. |
| `na.rm` | `FALSE` | all | Drop features with missing values after extraction/prep. |
| `folds` | `5` | all | Outer CV folds. |
| `seed` | `1234` | all | Reproducibility seed. |
| `base_learner` | `"SL.BART"` | all | Base learner. Use `SL.*` IDs for continuous/binary and explicitly set a supported `surv.*` ID for survival runs. |
| `verbose` | `FALSE` | all | Print progress. |
| `family` | `gaussian()` | all | Non-survival: `gaussian()`/`binomial()`. Survival is auto-detected from metadata or family. |
| `...` | — | all | Passed to the selected backend (`IL_conbin` or `ILsurv`). |

### Conbin-Specific Parameters (Continuous/Binary Path)

| Parameter | Default | Description |
|---|---|---|
| `base_screener` | `"All"` | SuperLearner screener used with base learner. |
| `meta_learner` | `"SL.nnls.auc"` | Stacked meta learner for late fusion. |
| `run_stacked` | `TRUE` | Enables late-fusion stacked model. |
| `run_concat` | `TRUE` | Enables early-fusion concatenated model. |
| `print_learner` | `TRUE` | Prints fit summary. |
| `refit.stack` | `FALSE` | Refit stacked learner on full data for final predictions. |

### Survival-Specific Parameters (via `...`)

| Parameter | Default | Description |
|---|---|---|
| `do_early_fusion` | `TRUE` | Train an early-fusion survival model on all features. |
| `weight_method` | `"IBS"` or `"COX"` | Late-fusion weighting objective. |
| `intermediate_learners` | `c("surv.coxph")` | Intermediate fusion learners on layer-risk features. |
| `t_vec`, `t_vec_probs` | `NULL`, quantiles | Time grid used in COX-style weighting summaries. |
| `layer_score` | `"sum"` | Aggregation of cumulative hazard increments (`sum`, `mean`, `l2`). |
| `weight_lambda` | `0.02` | Regularization strength for COX weighting optimizer. |
| `weight_penalty` | `"l2_to_uniform"` or `"entropy"` | Penalty used while learning survival late-fusion weights. |
| `weight_cap` | `1.0` | Optional cap on individual layer weights. |
| `optim_maxit_cox` | `4000` | Max iterations for COX weighting optimization. |
| `optim_maxit_ibs` | `300` | Max iterations for IBS weighting optimization. |
| `ibs_shrink_to_uniform` | `0` | Shrink IBS weights toward uniform blend. |

## Supported Models and Fusion Modules

### Supported Models

| Path | Supported base models |
|---|---|
| Continuous/Binary (`IL_conbin`) | Any `SuperLearner`-compatible `SL.*` learner available in your R session. Package wrappers include: `SL.BART`, `SL.LASSO`, `SL.enet`, `SL.glmnet2`, `SL.horseshoe`, `SL.mxBART` (plus standard SuperLearner learners such as `SL.glm`, `SL.randomForest`, etc.). |
| Survival (`ILsurv`) | Built-in survival learner IDs: `surv.rfsrc`, `surv.ranger`, `surv.coxboost`, `surv.bart`, `surv.coxph`, `surv.glmnet`. |

### Supported Fusion Outputs

| Path | Single-layer | Early fusion | Late fusion | Intermediate fusion |
|---|---|---|---|---|
| Continuous/Binary | Yes | `run_concat = TRUE` | `run_stacked = TRUE` with `meta_learner` | Not a separate branch in `IL_conbin` |
| Survival | Yes (`train_out$single`) | `do_early_fusion = TRUE` | Weighted layer blending (`weight_method = "IBS"` or `"COX"`) | `intermediate_learners` on layer-risk summaries |

## Output Reference: What You Get and How to Access It

This section summarizes the outputs produced by each integration method and where to find weights/importance values.

### Conbin Outputs (Binary/Continuous)

| Method | What it returns | Where to access |
|---|---|---|
| Single-layer (per omics layer) | Layer-specific predictions and metrics | `fit$yhat.train[, layer_name]`, `fit$yhat.test[, layer_name]` (if validation), `fit$AUC.train` / `fit$AUC.test` (binomial), `fit$R2.train` / `fit$R2.test` (gaussian) |
| Early fusion (concatenated) | One model on all features concatenated | Enable with `run_concat = TRUE`; outputs in `fit$yhat.train[, "concatenated"]`, `fit$model_fits$model_concat`, `fit$SL_fits$SL_fit_concat` |
| Late fusion (stacked) | Meta-model over layer-level predictions | Enable with `run_stacked = TRUE`; outputs in `fit$yhat.train[, "stacked"]`, `fit$model_fits$model_stacked`, `fit$SL_fits$SL_fit_stacked` |
| Layer weights (stacked) | Relative contribution of each layer in late fusion | `fit$weights` (available when `meta_learner = "SL.nnls.auc"` and `run_stacked = TRUE`) |

### Survival Outputs (Single/Early/Late/Intermediate)

| Method | Training outputs | Validation outputs |
|---|---|---|
| Single-layer | `fit$train_out$single$metrics` | `fit$valid_out$single$valid_cindex`, `fit$valid_out$single$valid_auc` |
| Early fusion | `fit$train_out$early$train_cindex`, `fit$train_out$early$train_auc` | `fit$valid_out$early$valid_cindex`, `fit$valid_out$early$valid_auc` |
| Late fusion | `fit$train_out$late$weights`, `fit$train_out$late$train_cindex`, `fit$train_out$late$train_auc` | `fit$valid_out$late$valid_cindex`, `fit$valid_out$late$valid_auc` |
| Intermediate fusion | `fit$train_out$intermediate` (per learner listed in `intermediate_learners`) | `fit$valid_out$intermediate` |

### Importance Outputs (Conbin and Survival)

| Importance type | Where to access | Notes |
|---|---|---|
| Conbin signed global feature importance | `fit$feature_importance_signed` | Always returned for non-survival fits; named numeric vector sorted by effect magnitude/sign. |
| Conbin signed per-layer importance | `fit$feature_importance_signed_by_layer` | List split by `featureType`. |
| Survival early-fusion combined importance | `fit$train_out$early$combined_importance` | Available when `do_early_fusion = TRUE`. |
| Survival late-fusion combined importance | `fit$train_out$late$combined_importance` | Weighted signed importance; names are prefixed like `layer::feature`. |
| BART-specific layer importance (optional) | `bartMachine::investigate_var_importance(fit$model_fits$model_layers[[layer]], plot = FALSE)` | Only for BART-based conbin fits (`base_learner = "SL.BART"`). |

### Quick Access Snippets

```{r, eval=FALSE}
# ---- Conbin: weights + top features ----
fit$weights
head(fit$feature_importance_signed, 20)
names(fit$feature_importance_signed_by_layer)
head(fit$feature_importance_signed_by_layer[[1]], 20)

# ---- Survival: late-fusion weights + top combined features ----
fit_surv$train_out$late$weights
head(fit_surv$train_out$late$combined_importance, 20)

# ---- Survival: inspect all fusion branches ----
fit_surv$train_out$single
fit_surv$train_out$early
fit_surv$train_out$late
fit_surv$train_out$intermediate
```

## Example 1: Binary Outcome (IBD Classification)

This section uses the PRISM dataset (Franzosa et al., 2019) for classifying IBD status. The binary target is coded in `sample_metadata$Y`.

### Step 1: Load and Inspect Training and Validation Data

```{r}
# Training data
load(url("https://github.com/himelmallick/IntegratedLearner/blob/master/data/PRISM.RData?raw=true"))

feature_table <- pcl$feature_table
sample_metadata <- pcl$sample_metadata
feature_metadata <- pcl$feature_metadata
rm(pcl)

# Quick checks
head(feature_table[1:5, 1:5])
head(sample_metadata[1:5, ])
head(feature_metadata[1:5, ])

table(feature_metadata$featureType)
table(sample_metadata$Y)

all(rownames(feature_table) == rownames(feature_metadata))
all(colnames(feature_table) == rownames(sample_metadata))

# Independent validation data
load(url("https://github.com/himelmallick/IntegratedLearner/blob/master/data/NLIBD.RData?raw=true"))
feature_table_valid <- pcl$feature_table
sample_metadata_valid <- pcl$sample_metadata
rm(pcl)

all(rownames(feature_table) == rownames(feature_table_valid))
all(colnames(feature_table_valid) == rownames(sample_metadata_valid))
```

### Step 2: Build PCL Inputs

```{r}
PCL_train <- list(
  feature_table = feature_table,
  sample_metadata = sample_metadata,
  feature_metadata = feature_metadata
)

PCL_valid <- list(
  feature_table = feature_table_valid,
  sample_metadata = sample_metadata_valid,
  feature_metadata = feature_metadata
)
```

### Step 3: Fit the Model

`IntegratedLearner` fits one model per layer (`base_learner`) and then combines layer-level predictions with a meta-learner (`meta_learner`).

```{r, warning=FALSE}
fit <- IntegratedLearner(
  PCL_train = PCL_train,
  PCL_valid = PCL_valid,
  folds = 5,
  base_learner = "SL.randomForest",
  meta_learner = "SL.nnls.auc",
  verbose = TRUE,
  family = binomial()
)
```

### Step 4: Inspect and Interpret Outputs

Core outputs for binary tasks include:

- `fit$AUC.train` and `fit$AUC.test`: AUC per layer and fusion model.
- `fit$weights`: layer contributions in the stacked model (when `SL.nnls.auc` is used).
- `fit$yhat.train` and `fit$yhat.test`: predicted probabilities.

```{r}
fit$AUC.train
fit$AUC.test
fit$weights
```

Plot ROC summaries for train and validation sets:

```{r, warning=FALSE, fig.height=10, fig.width=7}
plot.obj <- IntegratedLearner:::plot.learner(fit)
```

In this PRISM setting, you can compare which single layer is strongest and whether stacked fusion outperforms both individual layers and simple concatenation.

## Example 2: Continuous Outcome (Gestational Age)

This section uses the pregnancy dataset (Ghaemi et al., 2019), where `Y` is continuous gestational age.

### Step 1: Load and Inspect Data

```{r}
load(url("https://github.com/himelmallick/IntegratedLearner/blob/master/data/pregnancy.RData?raw=true"))

feature_table <- pcl$feature_table
sample_metadata <- pcl$sample_metadata
feature_metadata <- pcl$feature_metadata
rm(pcl)

head(feature_table[1:5, 1:5])
head(sample_metadata[1:5, ])
head(feature_metadata[1:5, ])

table(feature_metadata$featureType)
length(unique(sample_metadata$subjectID))

all(rownames(feature_table) == rownames(feature_metadata))
all(colnames(feature_table) == rownames(sample_metadata))

# Optional speed-up for local experimentation
# top_n <- 50
# subsetIDs <- c(1:top_n, (nrow(feature_table) - top_n + 1):nrow(feature_table))
# feature_table <- feature_table[subsetIDs, ]
# feature_metadata <- feature_metadata[subsetIDs, ]
```

### Step 2: Build PCL Input

```{r}
PCL_train <- list(
  feature_table = feature_table,
  sample_metadata = sample_metadata,
  feature_metadata = feature_metadata
)
```

### Step 3: Fit Continuous Model

For this example, we use BART base learners (`SL.BART`) with leave-one-subject-out style CV (`folds = 17`).

```{r, warning=FALSE}
fit <- IntegratedLearner(
  PCL_train = PCL_train,
  folds = 17,
  base_learner = "SL.BART",
  meta_learner = "SL.nnls.auc",
  family = gaussian()
)
```

### Step 4: Evaluate Predictive Accuracy

For continuous outcomes, `IntegratedLearner` reports `R2.train` (and `R2.test` if validation is provided).

```{r}
fit$R2.train
```

```{r fig.asp=0.8, fig.width=7}
plot.obj <- IntegratedLearner:::plot.learner(fit)
```

### Step 5: Uncertainty and Feature-Level Interpretation (BART)

When using `SL.BART`, you can inspect posterior predictive distributions and derive weighted posterior summaries.

```{r}
weights <- fit$weights

dataX <- fit$X_train_layers
dataY <- fit$Y_train

post.samples <- vector("list", length(weights))
names(post.samples) <- names(dataX)

for (i in seq_along(post.samples)) {
  post.samples[[i]] <- bart_machine_get_posterior(
    fit$model_fits$model_layers[[i]],
    dataX[[i]]
  )$y_hat_posterior_samples
}

weighted.post.samples <- Reduce("+", Map("*", post.samples, weights))
rownames(weighted.post.samples) <- rownames(dataX[[1]])
names(dataY) <- rownames(dataX[[1]])
```

Visualize 68% and 95% credible intervals for observations:

```{r fig.width=9, fig.height=7}
ord_names <- names(sort(rowMeans(weighted.post.samples), decreasing = TRUE))

mcmc_intervals(t(weighted.post.samples), prob = 0.68, prob_outer = 0.95) +
  scale_y_discrete(limits = ord_names) +
  geom_point(aes(x = dataY[ord_names], y = ord_names), shape = 1, size = 3, color = "black") +
  coord_flip() +
  theme_bw() +
  labs(
    x = "Gestational age (in months)",
    y = "Observations"
  ) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

Layer weights and feature-level inclusion proportions can also be examined for biological interpretation.

```{r}
omicsEye_theme <- function() {
  angle <- 45
  hjust <- 1
  ggplot2::theme_bw() +
    ggplot2::theme(
      axis.text.x = ggplot2::element_text(size = 8, vjust = 1, hjust = hjust, angle = angle),
      axis.text.y = ggplot2::element_text(size = 8, hjust = 1),
      axis.title = ggplot2::element_text(size = 10),
      plot.title = ggplot2::element_text(size = 10),
      plot.subtitle = ggplot2::element_text(size = 8),
      legend.title = ggplot2::element_text(size = 6, face = "bold"),
      legend.text = ggplot2::element_text(size = 7),
      axis.line = ggplot2::element_line(colour = "black", linewidth = 0.25),
      axis.line.x = ggplot2::element_line(colour = "black", linewidth = 0.25),
      axis.line.y = ggplot2::element_line(colour = "black", linewidth = 0.25),
      panel.border = ggplot2::element_blank(),
      panel.grid.major = ggplot2::element_blank(),
      panel.grid.minor = ggplot2::element_blank()
    )
}

safe_var_importance <- function(model, layer_label) {
  tryCatch({
    qq <- bartMachine::investigate_var_importance(model, plot = FALSE)
    df <- cbind.data.frame(qq$avg_var_props, qq$sd_var_props)
    colnames(df) <- c("mean", "sd")
    df$type <- layer_label
    df
  }, error = function(e) {
    warning(sprintf("Skipping variable importance for %s: %s", layer_label, conditionMessage(e)))
    data.frame(mean = numeric(), sd = numeric(), type = character())
  })
}

vimp_stack <- cbind.data.frame(fit$weights)
colnames(vimp_stack) <- "mean"
vimp_stack$sd <- NA
vimp_stack$type <- "stack"

layer_names <- names(fit$model_fits$model_layers)
vimp_layers <- lapply(layer_names, function(layer_nm) {
  safe_var_importance(fit$model_fits$model_layers[[layer_nm]], layer_nm)
})

vimp_layers <- vimp_layers[lengths(vimp_layers) > 0]
vimp_top <- do.call(
  rbind,
  lapply(vimp_layers, function(df) head(df[order(-df$mean), , drop = FALSE], 20))
)

VIMP <- as.data.frame(rbind.data.frame(vimp_stack, vimp_top))
VIMP <- tibble::rownames_to_column(VIMP, "ID")

p4 <- VIMP %>%
  dplyr::filter(type == "stack") %>%
  dplyr::arrange(desc(mean)) %>%
  ggplot(aes(y = mean, x = reorder(ID, -mean))) +
  geom_bar(stat = "identity", fill = "darkseagreen") +
  theme_bw() +
  omicsEye_theme() +
  ylab("Layer Weights") +
  xlab("")

p5 <- VIMP %>%
  dplyr::filter(type != "stack") %>%
  dplyr::arrange(mean) %>%
  dplyr::mutate(ID = stringr::str_replace_all(ID, stringr::fixed("_"), " ")) %>%
  ggplot(aes(reorder(ID, -mean), mean, fill = type)) +
  facet_wrap(. ~ type, scales = "free") +
  geom_bar(stat = "identity", fill = "lightsalmon") +
  geom_errorbar(aes(ymin = ifelse(mean - sd > 0, mean - sd, 0), ymax = mean + sd),
                width = 0.2,
                position = position_dodge(0.9)) +
  theme_bw() +
  coord_flip() +
  omicsEye_theme() +
  theme(strip.background = element_blank()) +
  ylab("Inclusion proportion") +
  xlab("")
```

```{r fig.width=7, fig.height=4}
plot_grid(
  p4,
  ncol = 1,
  labels = c("Estimated IntegratedLearner Layer Weights"),
  label_size = 8,
  vjust = 0.1
) + theme(plot.margin = unit(c(0.5, 0.5, 0.5, 0.5), "cm"))
```

```{r fig.width=9, fig.height=7}
plot_grid(
  p5,
  ncol = 1,
  labels = c("Top Features by Layer (BART Inclusion Proportions)"),
  label_size = 8,
  vjust = 0.1
) + theme(plot.margin = unit(c(0.5, 0.5, 0.5, 0.5), "cm"))
```

## Example 3: Survival Outcome (Time-to-event)

For survival tasks, `IntegratedLearner` dispatches to `ILsurv` when survival metadata are detected. The expected fields are:

- `time`: follow-up time (non-negative).
- `event`: event indicator (`0/1`).
- optional `Y`: `Surv(time, event)` convenience column.

This section provides a complete MAE workflow, followed by an equivalent PCL sketch.

### Step 1: Construct MAE Survival Inputs from TCGA_BRCA

```{r}


load(url("https://github.com/himelmallick/IntegratedLearner/blob/master/data/Survival/TCGA_BRCA.RData?raw=true"))
# Objects loaded: gene_all, mir_all

# Keep deterministic ordering
gene_all <- gene_all[order(gene_all$patient_id), ]
mir_all  <- mir_all[order(mir_all$patient_id), ]
rownames(gene_all) <- gene_all$patient_id
rownames(mir_all)  <- mir_all$patient_id

# Sample metadata
sample_ids <- rownames(gene_all)
sample_metadata <- data.frame(
  subjectID = sample_ids,
  time = gene_all$OS.time,
  event = gene_all$OS,
  Y = Surv(gene_all$OS.time, gene_all$OS),
  row.names = sample_ids
)

# Feature matrices (features in rows)
drop_cols <- c("patient_id", "age", "stage_i", "stage_ii", "race_white", "OS.time", "OS")
mat_gene <- t(as.data.frame(gene_all[, setdiff(names(gene_all), drop_cols)]))
mat_mirna <- t(as.data.frame(mir_all[, setdiff(names(mir_all), drop_cols)]))

# Per-layer SummarizedExperiments
se_gene <- SummarizedExperiment(
  assays = list(relative_abundance = mat_gene),
  colData = S4Vectors::DataFrame(sample_metadata[colnames(mat_gene), ])
)

se_mirna <- SummarizedExperiment(
  assays = list(pathway_abundance = mat_mirna),
  colData = S4Vectors::DataFrame(sample_metadata[colnames(mat_mirna), ])
)

mae <- MultiAssayExperiment(
  experiments = list(taxonomy = se_gene, pathway = se_mirna),
  colData = S4Vectors::DataFrame(sample_metadata)
)
```

### Step 2: Split MAE into Training and Validation

```{r}
set.seed(1)
idx <- sample(seq_along(sample_ids), size = floor(0.7 * length(sample_ids)))
mae_train <- mae[, idx, ]
mae_valid <- mae[, -idx, ]

# Utility to keep Y/time/event present in every experiment colData after subsetting
propagate_Y <- function(mae_obj) {
  exps <- MultiAssayExperiment::experiments(mae_obj)
  for (nm in names(exps)) {
    se <- exps[[nm]]
    if (!"Y" %in% colnames(SummarizedExperiment::colData(se))) {
      cd <- as.data.frame(MultiAssayExperiment::colData(mae_obj))[colnames(se), , drop = FALSE]
      SummarizedExperiment::colData(se) <- S4Vectors::DataFrame(cd)
    }
    exps[[nm]] <- se
  }
  MultiAssayExperiment::experiments(mae_obj) <- exps
  mae_obj
}

mae_train <- propagate_Y(mae_train)
mae_valid <- propagate_Y(mae_valid)
```

### Step 3: Fit Survival Model (MAE Mode)

```{r,warning=FALSE}
fit_surv_mae <- IntegratedLearner(
  MAE_train = mae_train,
  MAE_valid = mae_valid,
  experiment = c("taxonomy", "pathway"),
  assay.type = c("relative_abundance", "pathway_abundance"),
  folds = 2,
  base_learner = "surv.coxboost",  # alternatives: surv.coxph, surv.ranger, surv.glmnet, ...
  weight_method = "COX",           # alternative: "IBS"
  verbose = TRUE
)
```

### Step 4: Interpret Survival Outputs

```{r}
# Single-layer training metrics
# fit_surv_mae$train_out$single$metrics

# Late fusion (weighted integration)
fit_surv_mae$train_out$late$weights
fit_surv_mae$train_out$late$train_cindex
fit_surv_mae$train_out$late$train_auc

# Early and intermediate fusion summaries
#fit_surv_mae$train_out$early
#fit_surv_mae$train_out$intermediate

# Validation metrics
fit_surv_mae$valid_out$late$valid_cindex
fit_surv_mae$valid_out$late$valid_auc
fit_surv_mae$valid_out$single$valid_cindex
```

### Equivalent PCL Survival Pattern (Same Data)

```{r,warning =FALSE}
# Convert to PCL-style objects if preferred:
feature_table <- as.data.frame(rbind(mat_gene, mat_mirna))
feature_metadata <- rbind(
  data.frame(featureID = rownames(mat_gene), featureType = "gene", row.names = rownames(mat_gene)),
  data.frame(featureID = rownames(mat_mirna), featureType = "mirna", row.names = rownames(mat_mirna))
)

PCL_train <- list(
  feature_table = feature_table[, rownames(sample_metadata)[idx], drop = FALSE],
  sample_metadata = sample_metadata[rownames(sample_metadata)[idx], , drop = FALSE],
  feature_metadata = feature_metadata
)

PCL_valid <- list(
  feature_table = feature_table[, rownames(sample_metadata)[-idx], drop = FALSE],
  sample_metadata = sample_metadata[rownames(sample_metadata)[-idx], , drop = FALSE],
  feature_metadata = feature_metadata
)

fit_surv_pcl <- IntegratedLearner(
  PCL_train = PCL_train,
  PCL_valid = PCL_valid,
  folds = 2,
  base_learner = "surv.coxboost",
  weight_method = "IBS"
)
```

## Session Information

```{r}
sessionInfo()
```

## References

Franzosa EA et al. (2019). [Gut microbiome structure and metabolic activity in inflammatory bowel disease](https://www.ncbi.nlm.nih.gov/pubmed/30531976). *Nature Microbiology* 4(2):293-305.

Ghaemi MS et al. (2019). [Multiomics modeling of the immunome, transcriptome, microbiome, proteome and metabolome adaptations during human pregnancy](https://pubmed.ncbi.nlm.nih.gov/30561547/). *Bioinformatics* 35(1):95-103.

## Citation

Mallick et al. (2024). [An integrated Bayesian framework for multi-omics prediction and classification](https://pubmed.ncbi.nlm.nih.gov/38146838/). *Statistics in Medicine* 43(5):983-1002.
